### 2011-11-07 doc/pldi2012/experiments.tex

@@

\begin{figure}
  \centering
  \subfigure[Performance Scaling]{
    \begin{tikzpicture}
      \begin{axis}[xlabel=Number of Nodes,ylabel=Normalized Speedup,height=5cm,width=8.5cm,xtick={1,2,3,4}]
        \addplot coordinates {
% 12 CPUs
  (1, 5.29)
  (2, 7.34)
  (3, 7.30)
  (4, 10.14)
        };
        \addplot coordinates {
% 6 CPUs
  (1, 4.51)
  (2, 6.00)
  (3, 6.57)
  (4, 8.83)
        };
        \addplot coordinates {
% 4 CPUs
  (1, 3.36)
  (2, 4.05)
  (3, 5.86)
  (4, 6.68)
        };
        \addplot coordinates {
% 3 CPUs
  (1, 2.58)
  (2, 3.55)
  (3, 4.35)
  (4, 5.61)
        };
        \addplot coordinates {
% 2 CPUs
  (1, 1.83)
  (2, 2.43)
  (3, 3.48)
  (4, 4.15)
        };
        \addplot coordinates {
% 1 CPUs
  (1, 1.00)
  (2, 1.61)
  (3, 1.87)
  (4, 2.33)
        };
      \legend{12 CPU/node,6 CPU/node,4 CPU/node,3 CPU/node,2 CPU/node,1 CPU/node}
      \end{axis}
    \end{tikzpicture}
    \label{sfig:results2:perf}
  }

>>>>

\pgfplotsset{every axis legend/.append style={
  font=\tiny,
  at={(0.02,0.98)},
  anchor=north west},
  cycle list={{black, mark=x}, {black, mark=+},
              {black, mark=o}, {black, mark=diamond},
              {black, mark=triangle}, {black, mark=star}}};

@@

\usepgflibrary{plotmarks}
\begin{figure}
  \centering
  \subfigure[Performance Scaling]{
    \begin{tikzpicture}
      \begin{axis}[
          xlabel=Number of Nodes,
          ylabel=Normalized Speedup,
          height=5cm,
          width=8.5cm,
          xtick={1,2,3,4},
          legend columns=3,
        ]
        \addplot coordinates {
% 12 CPUs
  (1, 5.29)
  (2, 7.34)
  (3, 7.30)
  (4, 10.14)
        };
        \addplot coordinates {
% 6 CPUs
  (1, 4.51)
  (2, 6.00)
  (3, 6.57)
  (4, 8.83)
        };
        \addplot coordinates {
% 4 CPUs
  (1, 3.36)
  (2, 4.05)
  (3, 5.86)
  (4, 6.68)
        };
        \addplot coordinates {
% 3 CPUs
  (1, 2.58)
  (2, 3.55)
  (3, 4.35)
  (4, 5.61)
        };
        \addplot coordinates {
% 2 CPUs
  (1, 1.83)
  (2, 2.43)
  (3, 3.48)
  (4, 4.15)
        };
        \addplot coordinates {
% 1 CPUs
  (1, 1.00)
  (2, 1.61)
  (3, 1.87)
  (4, 2.33)
        };
      \legend{12 CPU/node,6 CPU/node,4 CPU/node,3 CPU/node,2 CPU/node,1 CPU/node}
      \end{axis}
    \end{tikzpicture}
    \label{sfig:results2:perf}
  }

### 2011-11-07 doc/pldi2012/highlevel.tex

\section{High-Level Runtime} 
\label{sec:highlevel}
In this Section we describe a high level runtime system
that is designed to implement the Legion programming
and execution models while remaining machine agnostic.  
The high level runtime system operates on top of the 
low-level runtime which hides machine specific details
and is described in detail in Section \ref{sec:lowlevel}.

The biggest challenge to implementing the high level
runtime is to avoid the latencies associated with distributed
memory systems and deep memory hierarchies.  To aid
in solving this problem, the high level runtime employs two useful 
scheduling techniques.  The first technique was proved
in Section \ref{sec:exec}: all scheduling decisions for a
task can be made locally in the context in which the
task was created.  The second technique is a {\em deferred execution model} 
that allows the runtime to hide the latencies associated
with moving data through deep memory hierarchies.  A
deffered execution model decouples the issuing of operations
from when they are actually performed.  This model is supported
by features in the low-level runtime that are discussed in Section \ref{sec:lowlevel}.

@@

A deffered execution model however presents challenges to implementing
the semantics of the Legion programming model.  Deffered execution
does not necessarily imply that tasks will be executed in the order in
which they are scheduled, but only in the order in which dependences
are expressed.  To orchestrate the execution of Legion
programs in this environment the high-level runtime is architected
in a manner similar to an out-of-order hardware processor, but at
a coarser granularity with tasks corresponding to instructions and
regions corresponding to registers.  There are multiple stages to 
executing a task which allows the runtime to extract
as much task-level parallelism as possible from an application whithin
the constraints of the Legion programming model.  To illustrate
the implementation of the high level runtime we now describe the
details of how each stage of task execution is performed.

\subsection{Dependence Analysis} 
\label{subsec:depanalysis}
The first step in the execution of a task is dependence analysis.  When
a task is registered with the high level runtime the task call includes
information about both the logical regions that the task will require
when it is executed as well as the read-write privileges and coherence properties
for each of the regions.  The runtime uses this information to perform
dependence detection with each of the previous tasks that have been registered
in the same context.  Dependence detection only needs to be performed at
the scope of an enclosing context as was proved in Section \ref{sec:exec}.

@@

\subsubsection{Dependence Detection}
\label{subsec:depdetect}
When a task is registered, the runtime
will perform dependence detection for every region that the task requests with
prior tasks that have been registered in the same context.  Since
tasks are registered in program order we will always perform dependence detection
in program order which guarantees correctness even if tasks are executed out-of-order.  
Performing dependence detection in program order is annalogous to how an out of order
processor works.  Out-of-order processors load instructions in order and only after
analyzing their instructions and register dependencies do they execute instructions
out of order.  The regions that a task requires are annalogous
to the registers listed in an instruction and the goal of the high level
runtime is to detect the same dependences and use them to enforce scheudling constraints
upon a task.  However, there is a catch to this analogy: in the Legion programming
model two different logical regions are not necessarily independent from each other.  In
order to perform dependence detection we rely on knowledge of the forest of regions to be
able to detect dependences between tasks which require different logical regions.

@@

\subsubsection{Managing Data Versions}
\label{subsec:dataflow}
One aspect of the deffered execution model is that it allows dependence analysis to run
ahead of the actual execution of tasks.  As a result the runtime must maintain information
about which {\em physical instances} of a logical region a task is allowed to use when executing.
A physical instance is a copy of the data in a logical region.  There can be multiple
physical instances of a logical region.  It is the responsibility of the high level runtime
to ensure tasks use instances corresponding to the appropriate version of data.

@@

These eight task calls give the programmer complete control over where all tasks are executed, and
where all data is placed within the memory hierarchy.  While this interface does give the programmer
complete control over data and task movement, it will never allow the programmer to affect the
correctness of the program.  The highlevel runtime still manages all the task dependences and the
ordering in which tasks are executed and copies are performed.  Therefore, regardless of choices
made by any mapper, the resulting program will always be executed correctly.

@@

The high level runtime also supports the use of multiple mappers in the same application.  For each task 
call the programmer can choose a mapper to be used for the runtime calls made inside that
task.  This allows for the creation of library code which provides its own mappers indpendent
from any application level mappers.

@@

\subsection{Execution}
\label{subsec:execution}
Once task mapping is complete the high level runtime performs the necessary operations
to issue the task to the low level runtime.  It first creates any new physical instances
for the regions which the task requires.  It then issues the necessary copies required
for the task to execute contignent on the tasks producing the source physical instances
have finished.  Then it issues the task itself to the low level processor contingent
on all the copies for the task haveing finished as well as the the set of tasks
on which the issuing task dependends having finished.  These dependences are all encoded
using event dependencies in the low level runtime.  Events are described
in detail in Section \ref{subsec:events}.  Continuing the analogy of an out-of-order
processor, the execution phase is equivalent to issuing an instruction to an
execution unit, and having the execution unit defer the instruction's execution until all its
operands are ready.  The low-level runtime contains processors that act as execution
units that will run a task once all of the events required for the task to run
have occurred.

>>>>

\section{High-Level Runtime} 
\label{sec:highlevel}
In this Section we describe a high-level runtime system
that is designed to implement the Legion programming
and execution models while remaining machine agnostic.  
The high-level runtime system operates on top of the 
low-level runtime which hides machine specific details
and is described in detail in Section \ref{sec:lowlevel}.

The biggest challenge to implementing the high-level
runtime is to avoid the latencies associated with distributed
memory systems and deep memory hierarchies.  To aid
in solving this problem, the high-level runtime employs two useful 
scheduling techniques.  The first technique was proved
in Section \ref{sec:exec}: all scheduling decisions for a
task can be made locally in the context in which the
task was created.  The second technique is a {\em deferred execution model} 
that allows the runtime to hide the latencies associated
with moving data through deep memory hierarchies.  A
deffered execution model decouples the issuing of operations
from when they are actually performed.  This model is supported
by features in the low-level runtime that are discussed in Section \ref{sec:lowlevel}.

@@

A deffered execution model however presents challenges to implementing
the semantics of the Legion programming model.  Deffered execution
does not necessarily imply that tasks will be executed in the order in
which they are scheduled, but only in the order in which dependences
are expressed.  To orchestrate the execution of Legion
programs in this environment the high-level runtime is architected
in a manner similar to an out-of-order hardware processor, but at
a coarser granularity with tasks corresponding to instructions and
regions corresponding to registers.  There are multiple stages to 
executing a task which allows the runtime to extract
as much task-level parallelism as possible from an application whithin
the constraints of the Legion programming model.  To illustrate
the implementation of the high-level runtime we now describe the
details of how each stage of task execution is performed.

\subsection{Dependence Analysis} 
\label{subsec:depanalysis}
The first step in the execution of a task is dependence analysis.  When
a task is registered with the high-level runtime the task call includes
information about both the logical regions that the task will require
when it is executed as well as the read-write privileges and coherence properties
for each of the regions.  The runtime uses this information to perform
dependence detection with each of the previous tasks that have been registered
in the same context.  Dependence detection only needs to be performed at
the scope of an enclosing context as was proved in Section \ref{sec:exec}.

@@

\subsubsection{Dependence Detection}
\label{subsec:depdetect}
When a task is registered, the runtime
will perform dependence detection for every region that the task requests with
prior tasks that have been registered in the same context.  Since
tasks are registered in program order we will always perform dependence detection
in program order which guarantees correctness even if tasks are executed out-of-order.  
Performing dependence detection in program order is annalogous to how an out of order
processor works.  Out-of-order processors load instructions in order and only after
analyzing their instructions and register dependencies do they execute instructions
out of order.  The regions that a task requires are annalogous
to the registers listed in an instruction and the goal of the high-level
runtime is to detect the same dependences and use them to enforce scheudling constraints
upon a task.  However, there is a catch to this analogy: in the Legion programming
model two different logical regions are not necessarily independent from each other.  In
order to perform dependence detection we rely on knowledge of the forest of regions to be
able to detect dependences between tasks which require different logical regions.

@@

\subsubsection{Managing Data Versions}
\label{subsec:dataflow}
One aspect of the deffered execution model is that it allows dependence analysis to run
ahead of the actual execution of tasks.  As a result the runtime must maintain information
about which {\em physical instances} of a logical region a task is allowed to use when executing.
A physical instance is a copy of the data in a logical region.  There can be multiple
physical instances of a logical region.  It is the responsibility of the high-level runtime
to ensure tasks use instances corresponding to the appropriate version of data.

@@

These eight task calls give the programmer complete control over where all tasks are executed, and
where all data is placed within the memory hierarchy.  While this interface does give the programmer
complete control over data and task movement, it will never allow the programmer to affect the
correctness of the program.  The high-level runtime still manages all the task dependences and the
ordering in which tasks are executed and copies are performed.  Therefore, regardless of choices
made by any mapper, the resulting program will always be executed correctly.

@@

The high-level runtime also supports the use of multiple mappers in the same application.  For each task 
call the programmer can choose a mapper to be used for the runtime calls made inside that
task.  This allows for the creation of library code which provides its own mappers indpendent
from any application level mappers.

@@

\subsection{Execution}
\label{subsec:execution}
Once task mapping is complete the high-level runtime performs the necessary operations
to issue the task to the low level runtime.  It first creates any new physical instances
for the regions which the task requires.  It then issues the necessary copies required
for the task to execute contignent on the tasks producing the source physical instances
have finished.  Then it issues the task itself to the low level processor contingent
on all the copies for the task haveing finished as well as the the set of tasks
on which the issuing task dependends having finished.  These dependences are all encoded
using event dependencies in the low level runtime.  Events are described
in detail in Section \ref{subsec:events}.  Continuing the analogy of an out-of-order
processor, the execution phase is equivalent to issuing an instruction to an
execution unit, and having the execution unit defer the instruction's execution until all its
operands are ready.  The low-level runtime contains processors that act as execution
units that will run a task once all of the events required for the task to run
have occurred.

### 2011-11-07 doc/pldi2012/lowlevel.tex
intro
More modification to the high level runtime section

\section{Low-Level Runtime}

The job of the low-level runtime is to provide a machine-independent layer
for the high-level runtime to sit on top of.  It needs to hide the
ideosynchracies of the underlying machine, operating system, and device
drivers.  At the same time, it needs to avoid getting too far away from the 
native capabilities of the hardware and software in order to obtain an 
efficient implementation.

The target machines for Legion include both clusters and GPU-equipped nodes,
and neither of these are able to efficiently provide a uniformly addressable
and accessible memory space across the system.  The magic of an
apparently-unified address space is left to the high-level runtime, and the
interface to the low-level runtime is done in terms of handles rather than 
pointers for runtime objects and relative offsets rather than absolute 
pointers for resource allocations.

The low-level runtime chooses to provide a ``deferred execution'' model rather
than the more common ``immediate'' model.  Requests made of the low-level 
runtime usually return to the caller immediately, with the requested action
performing asynchronously to the caller's thread of execution.  This is done
for two different, but equally important, reasons.  First, it matches the
standard GPU execution model.  GPUs tend to have very deep pipelines, and
maximum performance cannot be achieved if each command has to be run to
completion before the next command can be started.  Second, in clusters, the
latency of inter-node communication can be significant, and a caller that
waited for every request of a remote node to return before continuing on would
also be similarly underutilizing the computational resources of the system.

Although it does not show up explicitly in the interface, the low-level runtime
was also architected to permit a distributed and hierarchical implementation.
Critical resources can be replicated or migrated between nodes to reduce
both average latencies and the amount of traffic that use the scarce inter-node
communication bandwidth.

\subsection{Events}

The basic building block of the low-level runtime's deferred execution model
is the Event.  An Event is a placeholder for the completion of a deferred
operation.  When a call is made to the low-level runtime, an Event is usually
returned to the caller.  The caller may query whether the event has occurred,
and may also ask to wait until the event occurs, but the more common case is
to provide one or more Events to subsequent requests made of the runtime, 
causing those requests to be automatically deferred until the previous
operations corresponding to those Events have completed.  These requests will
also return Events, and the caller can set up a complicated, but coordinated,
set of operations to be performed by the low-level runtime with no babysitting
required of the original caller.

With nearly every single runtime operation creating a new Event and referring
to potentially any number of past Events, both the time and space overhead of
these Events is a serious concern.  The time overhead is addressed by a
mechanism that allows a node to create a new Event without needing to
communicate with any other node.  The Event's handle identifies which node
created it, allowing another node to lazily access it only when and if it 
is asked to perform an operation on it.  Checking to see if an Event has
occurred is done via a subscription mechanism.  Any number of threads or
deferred operations on one node can query or wait on an Event with a single
message sent to the originating node.  The requesting node tracks the multiple
local requestors and upon receiving the trigger message from the originating
node, distributes that message to all the local parties that are interested.

The potential space concern with Events is addressed by a generation scheme
that allows Events to be reclaimed and reused without tracking references.  The
naive implementation of an Event cannot reclaim the storage used by an Event
until the Event has triggered \emph{and} all waiters on the event have been
notified.  Some of these waiters may take quite a while to make their query,
and worse, we must somehow track how many queries to expect if we want to know
when they've all been performed.  However, if the Event handle is paired with a
``generation'' count that tracks how many times the Event has been triggered,
the Event can be reused as soon as the previous generation has been triggered.
Holders of an older (handle,generation) pair are able to query the newer
incarnation of the Event and determine from the generation mismatch that their
Event must have already triggered.  With this technique, there is no need to
count references to the Event at all.  An Event can only be reused until the
generation counter reaches its maximum value, the current implementation
supports $2^{32}-1$ generations, and should that not be sufficient, extending
it to $2^{64}-1$ is trivial.

\subsection{Deferred Locks}

Events are used to provide a deterministic ordering between operations, but in
some cases, enforcing only serializability is a sufficient for correctness and
can result in much better performance.  The traditional tool for this job is
the mutex (a.k.a. lock), and it is from the common reader/writer lock that
the low-level runtime's \emph{deferred lock} is derived.  The key visible 
difference between a mutex and a deferred lock is that the request does not
block the caller's execution.  Many libraries offer ``nonblocking locks'', but
they achieve this by returning without the lock if it is unavailable.  If
a deferred lock is currently unavailable, the request is guaranteed to be
granted at some time in the future (with no further action required of the 
caller).  That time in the future is described with an Event.  Like all other
Events, it can be given to other threads, or to the runtime, or even stored
in a data structure in memory to be extracted by an arbitrary thread at an 
arbitrary time.  This dissociation of a held lock from the thread that
requested it is another important difference between the deferred lock and
existing mutexes.

The deferred nature of a deferred lock also allows a distributed implementation
of locks.  Similar to Events, a deferred lock can be created on any node
without communication with the other nodes in the cluster.  That node becomes
the initial home for the lock, but the deferred lock (unlike events) can 
migrate to other nodes or even be shared between nodes, allowing repeated
lock requests from a single node to be kept local to that node after a single
message is sent to the current owner.  (Should the ownership of the lock change
while the message is in flight, the previous owner forwards the request to
the new owner.  The highly unlikely, but possible, case in which multiple 
forwardings are required is bounded at the number of nodes in the cluster.)
The ability to migrate locks between nodes allows the performance of a
heavily-contended cluster-level lock to approach that of a single-node lock.

@@

\subsection{Processors}

Hardware units capable of performing computation are captured as generic
processors in the low-level runtime.  Like other runtime objects, a processor
has a globally unique name and can be referred to anywhere in the system.
Processors come in different \emph{kinds} (e.g. X86 CPU core, CUDA-capable GPU),
but they all support the same
interface and execution model.  A request to spawn a task on a given processor
may come from anywhere in the system.  The request names which task should
be spawn and provides input arguments and zero or more events.  Once those
events have occurred, the task is placed into a queue of ready tasks, which
the processor works through in the order in which they were enqueued.  A
processor will generally run a task to completion before taking another, but
will try to run another task (resources permitting) if the current task needs
to wait for an event to occur.  Once a task is placed on a processor's queue,
it may not be moved or deleted.  Load-balancing of tasks is expected to be done
by the high-level runtime.  The primary purpose of the queue for the low-level
Processor is to be able to hide the latency of starting new tasks.

\subsection{Memories}

Application-visible storage in the low-level runtime is described in terms of
memory objects.  A memory object has a size and describes its affinity to 
processors (and other memories) in terms of relative bandwidth and latency.
The low-level runtime allows for the allocation and freeing of region instances
and allocators.  It also supports bulk transfers from one memory to another.
These bulk transfers can often be implemented using dedicated DMA (or RDMA)
hardware, freeing up the processors to perform more interesting
computation.

\subsection{Regions}

Application-level regions are broken into four inter-related pieces in the
low-level runtime:
\begin{description}

\item[MetaData] The metadata for a logical region captures attributes of a
logical region that are common across all instances.  These attributes include
the size of the logical region, the element type, and the allocation mask,
which tracks which portions of the logical region are actually in use.  Region
MetaDatas are created independently of any memory objects and are accessible
(via a globally unique name) to all nodes in the machine.

@@

\item[Instance] Each physical instance of a region stores some version of all
the elements in a region.  As with allocators, a physical instance is
associated with a particular region metadata and spends its entire lifetime in
the memory objecy in which it was created.  When multiple physical instances
exist for the same region, the low-level runtime does nothing to keep them
coherent - it is the job of the high-level runtime to perform the appropriate
copies and set up the right inter-operation dependencies to provide the level
of coherence requested by the application.  Although a physical instance 
stores the contents of a region, it does not directly offer access to those
contents.

@@

\subsection{Initialization}

On startup, the low-level runtime allocates all the resources it needs from
the operating system and set up the communication channels between nodes.
Each processor in the system then automatically runs the high-level runtime's
initialization task, which will usually spawn more tasks.  These tasks are
started, and whenever any processor's task queue is empty , the high-level
runtime's scheduler task is automatically called to explore the possibility
of moving tasks around for load balancing.

\subsection{SMP Implementation}

The SMP implementation of the low-level runtime is a relatively simple one.
It runs on any SMP computer that supports POSIX threads.  A single memory
object is created for the shared memory, and a processor object is created for
each CPU core in the system.  Any GPUs in the system are ignore.  Events are 
mutex-protected data structures that maintain a list of callbacks to call when
the corresponding event is triggered.  Locks are similar, except that the list
is of pending lock requests to attempt when the previous lock is released.

\subsection{GPU Cluster Implementation}

The cluster implementation of the low-level runtime is implemented on a mix
of POSIX threads for intra-node threading and communication, GASnet for 
inter-node communication, and CUDA for access to resources of the GPUs in 
each node.  Locks and events have roughly the same implementation as the SMP
runtime for handling multiple local requestors, but use GASnet active messages
to coordinate lock ownership and event subscription and triggering between
nodes.  In addition to the processor objects for each CPU core, a single
processor object is created for each GPU in the system.  It would be preferable
to expose the GPU's computational resources at a finer granularity, but
current GPU hardware doesn't directly allow this flexibility and recent attempts
at implementing such control in a software layer on top of CUDA have suffered
from significant overhead.  The memory hierarchy is expanded to include a 
``global'' memory that is striped across all nodes and accessible via GASnet
RDMA calls, and two memories per GPU: one for the GPU's framebuffer that is
accessible only to that GPU and by bulk DMA operations, and one for the pinned
``zero copy'' system memory that is accessible to the GPU and all the CPU cores
in a single system.

>>>>

\section{Low-Level Runtime}
\label{sec:lowlevel}

The purpose of the low-level runtime is to provide a machine-independent layer
for the high-level runtime.  This abstraction will provide portability to the Legion
programming model and the high-level runtime.  To achieve this goal the low-level
runtime needs to hide the ideosynchracies of the underlying machine.  At the same time, it needs to avoid 
straying too far from the native capabilities of the underlying hardware 
and software in order to obtain an efficient implementation.  The Legion
low-level runtime does this by abstracting the machine into a base set
of objects required to implement the Legion high-level.  These objects,
include processors, memories, regions, events, and deffered locks.

%The target machines for Legion include both clusters and GPU-equipped nodes,
%and neither of which are able to efficiently provide a uniformly addressable
%and accessible memory space across the system.  The low-level runtime does
%not hide this fact from the high-level runtime, but rather relies on the
%high-level runtime to manage data movement.  Consequently all objects in
%the low-level runtime are named by handles 
%which can be copied by value and moved around the machine.

%The low-level runtime provides a deferred execution model rather
%than the more common immediate model where operations occur as soon
%as requested.  Requests made of the low-level 
%runtime usually return to the caller immediately, with the requested action
%performing asynchronously to the caller's thread of execution.  This is done
%for two different, but equally important, reasons.  First, it matches the
%standard GPU execution model.  GPUs tend to have very deep pipelines, and
%maximum performance cannot be achieved if each command has to be run to
%completion before the next command can be started.  Second, in clusters, the
%latency of inter-node communication can be significant, and a caller that
%waited for every request of a remote node to return before continuing on would
%also be similarly underutilizing the computational resources of the system.

%Although it does not show up explicitly in the interface, the low-level runtime
%was also architected to permit a distributed and hierarchical implementation.
%Critical resources can be replicated or migrated between nodes to reduce
%both average latencies and the amount of traffic that use the scarce inter-node
%communication bandwidth.

\subsection{Low-Level Abstractions}
\label{subsec:lowobjects}
\subsubsection{Events}
\label{subsec:events}
The low-level runtime supports a deffered execution model to help hide
the latencies associated with communicating in a distributed memory
system.  The basic building block of the low-level runtime's deferred execution model
is the {\em event}.  An event is a placeholder for the completion of a deferred
operation.  When an asynchronous call is made to any of the low-level runtime objects, an event is 
immediately returned to the caller.  The caller may query whether the event has {\em triggered}
which happens when an even occurs.  The common use case for events is
to provide them as prerequisites to subsequent requests made of the runtime. 
These dependences cause the later requests to be automatically deferred until the previous
operations have completed.  By using events as dependences, the caller can
set up complicated, but coordinated chains of operations to be executed without
having to wait for the low-level runtime to execute them.
%These requests will
%also return events, and the caller can set up a complicated, but coordinated,
%set of operations to be performed by the low-level runtime with no babysitting
%required of the original caller.

With nearly every single runtime operation creating a new event and referring
to potentially any number of past events, both the time and space overhead of
implementing events is a serious concern.  The time overhead is addressed by a
mechanism that allows a node to create a new event without needing to
communicate with any other node.  The event's handle identifies which node
created it, allowing another node to lazily access it only when and if it 
is asked to perform an operation on it.  Checking to see if an event has
occurred is done via a subscription mechanism.  Any number of threads or
deferred operations on one node can query or wait on an event with a single
message sent to the originating node.  The requesting node tracks the multiple
local requestors and upon receiving the trigger message from the originating
node, distributes that message to all the local parties that are interested.

The potential space concern with events is addressed by a {\em generation scheme}
that allows events to be reclaimed and reused without tracking references.  The
naive implementation of an event cannot reclaim the storage used by an event
until the event has triggered and all waiters on the event have been
notified.  Some of these waiters may take quite a while to query the event.
Additionaly, we must somehow track how many queries to expect to know
when they've all been performed.  However, if the event handle is paired with a
{\em generation count} that tracks how many times the event has been triggered,
the event can be reused as soon as the previous generation has been triggered.
Holders of an older (handle,generation) pair are able to query the newer
incarnation of the event and determine from the generation mismatch that their
event must have already triggered.  With this technique, there is no need to
count references to the event at all. 
% An event can only be reused until the
%generation counter reaches its maximum value, the current implementation
%supports $2^{32}-1$ generations, and should that not be sufficient, extending
%it to $2^{64}-1$ is trivial.

\subsubsection{Deferred Locks}
\label{subsec:defferedlocks}
Events are used to provide a deterministic ordering between operations, but in
some cases is overly restrictive.  To implement the more relaxed coherence
properties supported in Legion we need a less restrictive ordering semantics that
can result in better performance.  The traditional tool for this job is
the mutex (a.k.a. lock), and it is from the common reader/writer lock that
the low-level runtime's {\em deferred lock} is derived.  The key visible 
difference between a mutex and a deferred lock is that the request does not
block the caller's execution.  Many libraries offer nonblocking locks, but
they achieve this by returning without the lock if it is unavailable.  If
a deferred lock is currently unavailable, the request is guaranteed to be
granted at some time in the future (with no further action required of the 
caller).  That time in the future is described with an event.  Like all other
events, it can be given to other threads, or even stored
in a data structure in memory to be extracted by an arbitrary thread at a 
later time.  This dissociation of a held lock from the thread that
requested it is another important difference between the deferred lock and
existing mutexes.

The deferred nature of a deferred lock also allows a distributed implementation
of locks.  Similar to events, a deferred lock can be created on any node
without communication with the other nodes in the cluster.  That node becomes
the initial home for the lock, but the deferred lock (unlike events) can 
migrate to other nodes or even be shared between nodes, allowing repeated
lock requests from a single node to be kept local to that node after a single
message is sent to the current owner.  (Should the ownership of the lock change
while the message is in flight, the previous owner forwards the request to
the new owner.  The highly unlikely, but possible, case in which multiple 
forwardings are required is bounded at the number of nodes in the cluster.)
The ability to migrate locks between nodes allows the performance of a
heavily-contended cluster-level lock to approach that of a single-node lock.

@@

\subsubsection{Processors}
\label{subsec:processors}
Hardware units capable of performing computation are captured as generic
processors in the low-level runtime.  Like other runtime objects, a processor
has a globally unique name and can be referred to anywhere in the system.
Processors come in different \emph{kinds} (e.g. X86 CPU core, CUDA-capable GPU),
but they all support the same
interface and execution model.  A request to spawn a task on a given processor
may come from anywhere in the system.  The request names which task should
be spawned and provides input arguments and zero or more events.  Once those
events have occurred, the task is placed into a queue of ready tasks, which
the processor works through in the order in which they were enqueued.  A
processor will generally run a task to completion before taking another, but
will try to run another task (resources permitting) if the current task needs
to wait for an event to occur.  Once a task is placed on a processor's queue,
it may not be moved or deleted.  Load-balancing of tasks is expected to be done
by the high-level runtime.  The primary purpose of the queue for the low-level
Processor is to be able to hide the latency of starting new tasks.

\subsubsection{Memories}
\label{subsec:memories}
Application-visible storage in the low-level runtime is described in terms of
memory objects.  A memory object has a size and describes its affinity to 
processors (and other memories) in terms of relative bandwidth and latency.
The low-level runtime allows for the allocation and freeing of region instances
and allocators.  It also supports bulk transfers from one memory to another.
These bulk transfers can often be implemented using dedicated DMA (or RDMA)
hardware, freeing up the processors to perform more interesting
computation.

\subsubsection{Region Meta-Data Objects}
\label{subsec:regionmeta}
Application-level regions are broken into four inter-related pieces in the
low-level runtime:
\begin{description}

\item[MetaData] The metadata for a logical region captures attributes of a
logical region that are common across all instances.  These attributes include
the size of the logical region, the element type, and the allocation mask,
which tracks which portions of the logical region are actually in use.  
%Region
%MetaDatas are created independently of any memory objects and are accessible
%(via a globally unique name) to all nodes in the machine.

@@

\item[Instance] Each physical instance of a region stores some version of all
the elements in a region.  As with allocators, a physical instance is
associated with a particular region metadata and spends its entire lifetime in
the memory objecy in which it was created.  
%When multiple physical instances
%exist for the same region, the low-level runtime does nothing to keep them
%copies and set up the right inter-operation dependencies to provide the level
%of coherence requested by the application.  Although a physical instance 
%stores the contents of a region, it does not directly offer access to those
%contents.

@@

\subsection{Low-Level Implementations}
\label{subsec:lowimpl}
The primary goal of the low-level runtime is to provide an
abstraction that enables efficient execution of Legion programs
on a wide array of machines.  To support this claim we describe
two implementation of the low-level runtime: one based on POSIX
threads for a shared memory SMP and one based on GASNet and CUDA
for distributed memory clusters with multiple GPUs.
%The magic of an
%apparently-unified address space is left to the high-level runtime, and the
%interface to the low-level runtime is done in terms of handles rather than 
%pointers for runtime objects and relative offsets rather than absolute 
%pointers for resource allocations.

%\subsection{Initialization}

%On startup, the low-level runtime allocates all the resources it needs from
%the operating system and set up the communication channels between nodes.
%Each processor in the system then automatically runs the high-level runtime's
%initialization task, which will usually spawn more tasks.  These tasks are
%started, and whenever any processor's task queue is empty , the high-level
%runtime's scheduler task is automatically called to explore the possibility
%of moving tasks around for load balancing.

\subsubsection{SMP Implementation}
\label{subsec:smpimpl}
The SMP implementation of the low-level runs on any SMP computer that supports 
POSIX threads.  A single memory object is created for the shared memory, and 
a processor object is created for
each CPU core in the system.  Events are 
mutex-protected data structures that maintain a list of callbacks to call when
the corresponding event is triggered.  Locks are similar, except that the list
is of pending lock requests to attempt when the previous lock is released.

\subsubsection{GPU Cluster Implementation}
\label{subsec:clusterimpl}
The cluster implementation of the low-level runtime is implemented on a mix
of POSIX threads for intra-node threading and communication, GASnet for 
inter-node communication, and CUDA for access to resources of the GPUs in 
each node.  Locks and events have roughly the same implementation as the SMP
runtime for handling multiple local requestors, but use GASnet active messages
to coordinate lock ownership and event subscription and triggering between
nodes.  In addition to the processor objects for each CPU core, a single
processor object is created for each GPU in the system.  It would be preferable
to expose the GPU's computational resources at a finer granularity, but
current GPU hardware doesn't directly allow this flexibility and recent attempts
at implementing such control in a software layer on top of CUDA have suffered
from significant overhead.  The memory hierarchy is expanded to include a 
{\em global} memory that is striped across all nodes and accessible via GASnet
RDMA calls, one process local memory per cluster node, and two memories per GPU: one for the GPU's framebuffer that is
accessible only to that GPU and by bulk DMA operations, and one for the pinned
{\em zero copy} system memory that is accessible to the GPU and all the CPU cores
in a single system.
